---
title: "Working with Spatial Data in R"
author: "Patrick Sherry"
date: 05/11/2026
format: html
---

## Data

Alaska regional boundaries: Kibele, J. and J. Clark. 2018. State of Alaska's Salmon and People Regional Boundaries. Knowledge Network for Biocomplexity. <https://doi.org/10.5063/F1125QWP>

Community locations and population: Clark, J., S. Ochs, D. Strong, and National Historic Geographic Information System. 2018. Languages used in Alaskan households, 1990-2015. Knowledge Network for Biocomplexity. <https://doi.org/10.5063/F11G0JHX>


```{r}
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
```

```{r}
# read in shapefile using read_sf()
ak_rgns_sf <- read_sf(here("data/ak_regions_simp.shp"))

# read in population data
pop_df <- read_csv(here("data/alaska_population.csv"))

pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F)

ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

st_crs(ak_rgns_3338_sf)


pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)
pop_joined_sf <- st_join(pop_3338_sf, 
                         ak_rgns_3338_sf, 
                         join = st_within)

pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%
    group_by(region) %>%
    summarize(total_pop = sum(population))

pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf, 
                             pop_rgn_df, 
                             by = "region")

write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.shp"))


rivers_3338_sf <- read_sf(here("data/ak_rivers_simp.shp"))

pop_3857_sf <- st_transform(pop_3338_sf, 
                         crs = 3857)

epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

pop_rgn_4326_sf <- pop_rgn_3338_sf %>% 
    st_transform(crs = 4326)

pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_sf$city, ", population ", comma(pop_4326_sf$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m



```
## a08 – Spatial Data Analysis

### Decision 1: CRS 3338 (Alaska Albers) for the spatial join

I transformed both the population points and the Alaska regions shapefile to EPSG:3338 before running `st_join()`.
The spatial join requires both layers to share the same CRS, and Alaska Albers is the standard projected CRS for
Alaska because it minimizes area distortion across the state's wide east-west. Using a projected CRS rather than a
geographic CRS like WGS84 makes distance and area calculations more accurate.

### Decision 2: `st_within` vs `st_intersects` for assigning cities to regions

I used `join = st_within` so that each city point is assigned only to the region it is completely inside.
`st_intersects`, would also match points that are on a region boundary, which could assign a city to two regions
and misrepresent population totals for border areas. Because city locations are points, `st_within` is the more
precise.

### Decision 3: `ggspatial` basemap

I included `annotation_map_tile()` from `ggspatial` to provide geographic context behind the static ggplot2 map.
For the regional-level Alaska map, the basemap helps orient the reader and confirms that the region aligns with
real geography. For the interactive Leaflet version, the basemap is redundant since Leaflet's default tiles so I
omitted it there.